---
title: "APEC 8221 - Assignment 0: Workspace Setup"
subtitle: "Submission File"
author: "[Yunshu Yang]"
format: 
  pdf:
    colorlinks: true
execute:
  warning: false
  message: false
editor: visual
---

## Part 1: Software Installation & Verification {#sec-setup}

Below are the results from my run of the `setup-tests_autosave.R` script.

<!-- Do not change the code in the chunk below. -->

<!-- It reads and displays the contents of the `assignment0-setup.txt` file you created. -->

```{r}
#| label: verification-results
#| echo: false

# This chunk reads the text file you created in Part 1 and prints its contents here.
# Make sure the file path is correct. If your .qmd file and .txt file are in the
# same folder, this will work.
readLines("assignment0-setup.txt") |> 
  cat(sep = "\n")
```

------------------------------------------------------------------------

## Part 2: Git & GitHub Repository {#sec-git}

<!-- Make sure the file path correctly links to your repo URL. -->

My GitHub repo URL is <https://github.com/yunshuyang1129/apec8221-yang9291.git>

------------------------------------------------------------------------

## Part 3: Embedded Quarto Document {#sec-quarto}

This section embeds the content from the `assignment-0-quarto.qmd` file you created.

<!-- The line below includes your other Quarto file's content. -->

<!-- Make sure the file path is correct relative to this document. -->

{{< include assignment-0-quarto.qmd >}}

------------------------------------------------------------------------

## Part 4: R Basics Knowledge Check {#sec-r-practice}

### 1. Objects and Assignment

**Question:** What is the primary assignment operator in R? Use it to create an object named `my_age` that holds your age in years.

**Answer:** \[Your text answer here...\]

```{r}
#| label: q1-assignment

# Create the 'my_age' object below
my_age <- 24   
my_age
```

### 2. Vectors

**Question:** What is the purpose of the `c()` function in R? Use it to create a vector named `favorite_numbers` containing three of your favorite numbers.

**Answer:** \[Your text answer here...\]

```{r}
#| label: q2-vectors

# Create the 'favorite_numbers' vector below
favorite_numbers <- c(7, 13, 21)
favorite_numbers
```

### 3. Subsetting Vectors

**Question:** Write the code that would select the *second* number from the `favorite_numbers` vector you just created.

```{r}
#| label: q3-subsetting

# Write your subsetting code below
favorite_numbers[2]
```

### 4. Missing Values and Logic

**Question:** The code chunk below creates a vector `x` with missing values (`NA`). a. Write a line of code that returns a logical vector (`TRUE`/`FALSE`) indicating which elements are missing. b. Write a line of code that *counts* how many total missing values are in the vector.

```{r}
#| label: q4-setup
# This vector is for the exercise
x <- c(1, 5, NA, 10, NA, 20)
```

```{r}
#| label: q4a-find-na
# a) Return a logical vector indicating missing elements
is.na(x)
```

```{r}
#| label: q4b-count-na
# b) Count the total number of missing values
sum(is.na(x))
```

### 5. Data Frames and Inspection

**Question:** The code chunk below creates a simple data frame called `pets`. a. Use a function to look at the first few rows of the data frame. b. Use a function to get a summary of the structure of the data frame. c. Write the code to select *only* the `age` column from the data frame.

```{r}
#| label: q5-setup
# This data frame is for the exercise
pets <- data.frame(
  name = c("Fido", "Whiskers", "Nemo"), 
  animal = c("Dog", "Cat", "Fish"), 
  age = c(5, 8, 2)
)
```

```{r}
#| label: q5a-head
# a) Look at the first few rows
head(pets)
```

```{r}
#| label: q5b-structure
# b) Get a summary of the structure
head(pets)
```

```{r}
#| label: q5c-select-column
# c) Select only the 'age' column
pets$age
```

### 6. Workspace and Files

**Question:** a. What function would you use to see a list of all the objects currently in your R environment? b. What function would you use to find out your current working directory?

**Answer:** a. \[ls()\] b. \[getwd()\]

### 7. Getting Help

**Question:** Which function in R can you use to open the help documentation for another function? In the code chunk below, write the code you would use to see the help file for the `mean()` function.

**Answer:** \[?function_name or help("function_name")\]

```{r}
#| label: q7-help 
#| #| eval: false

# Write the code to see the help file for the mean() function below.
# I've set eval: false so this chunk won't cause an error when you render.
?mean
help("mean")
```
