---
title: "APEC 8221 - Assignment 1: Tidyverse Fundamentals"
author: "[REPLACE WITH YOUR NAME]"
date: today
format: 
  pdf:
    number-sections: true
    colorlinks: true
    embed-resources: true
execute:
  warning: false
  message: false
  eval: true
  echo: true
editor: visual
---

```{=html}
<!-- 
Before you begin:
1. Save this file as assignment1-[your-x500-id].qmd
2. Replace "[REPLACE WITH YOUR NAME]" above with your actual name
3. Make sure you're working in your course R Project
4. Include your GitHub repository URL in Canvas submission
-->
```

**GitHub Repository URL:** \[Insert your repository URL here\]

# Setup

```{r}
#| label: setup

# Load required packages
library(tidyverse)
library(gapminder)

# Check data structure
glimpse(gapminder)
```

# Data Analysis & Visualization Practice

## Top Performers Identification

Find the top 3 countries in each continent for highest life expectancy and highest GDP per capita in 2007.

### Highest Life Expectancy by Continent (2007)

```{r}
#| label: top-life-exp

# Your code here - find top 3 countries by life expectancy in each continent
# Remember to use group_by(), arrange(), and slice_head()

```

### Highest GDP per Capita by Continent (2007)

```{r}
#| label: top-gdp

# Your code here - find top 3 countries by GDP per capita in each continent

```

**Brief analysis:** *\[Write 2-3 sentences about what patterns you notice in the top performers\]*

## Development Progress Visualization

Create a scatter plot comparing each country's 1952 vs. 2007 life expectancy.

```{r}
#| label: life-exp-data-prep

# Step 1: Create separate data frames for 1952 and 2007
lifeExp_1952 <- gapminder %>%
  # Your code here - filter for 1952 and select relevant columns

lifeExp_2007 <- gapminder %>%
  # Your code here - filter for 2007 and select relevant columns (include population!)

# Step 2: Combine using provided left_join code
lifeExp_data <- left_join(lifeExp_1952, lifeExp_2007, by = "country")

# Step 3: Handle missing data
lifeExp_data <- lifeExp_data %>%
  # Your code here - remove countries with missing data using filter()

```

```{r}
#| label: life-exp-plot
#| fig-width: 10
#| fig-height: 8

# Create the scatter plot
ggplot(lifeExp_data, aes(x = lifeExp_1952, y = lifeExp_2007)) +
  # Your code here - add the diagonal reference line using geom_abline()
  # Add points with color by continent and size by population
  # Add professional labels and theme

```

**Analysis:** *\[Write 2-3 sentences explaining what the plot shows about global health progress\]*

## Development Progress Classification

Using the `lifeExp_data` from Task 2.1.2, classify countries by improvement level.

```{r}
#| label: life-exp-classification

# Add classification to your existing lifeExp_data
lifeExp_data <- lifeExp_data %>%
  mutate(
    improvement_level = case_when(
      # Your code here - use the provided classification logic
      # Major Improvement: 30+ years gained
      # Moderate Improvement: 15-29 years gained  
      # Minor Improvement: 0-14 years gained
      # Decline: Lost life expectancy
    )
  )

# Create summary table
improvement_summary <- lifeExp_data %>%
  # Your code here - group by continent and improvement_level, count countries

print(improvement_summary)
```

### Analysis Questions

**1. Which continent has the most countries with "Major Improvement"?**

```{r}
#| label: major-improvement-analysis

# Your code here - analyze major improvement by continent

```

\[Your answer here\]

**2. Are there any countries that experienced decline? Which ones and why might this have happened?**

```{r}
#| label: decline-analysis

# Your code here - find and examine countries with decline

```

*\[Your answer here\]*

**3. What does this tell us about global health convergence?**

*\[Your answer here - 2-3 sentences about convergence patterns\]*

## Economic Development Trajectories

Create a line plot showing life expectancy changes over time for selected countries.

### Step 1: Guided Discovery

```{r}
#| label: trajectory-discovery

# Find interesting countries for your plot
life_exp_changes <- lifeExp_data %>%
  # Your code here - calculate change and identify top/bottom performers

# Print top gainers and any decliners
# Your code here

```

### Step 2: Create Trajectory Visualization

```{r}
#| label: trajectory-plot
#| fig-width: 10
#| fig-height: 6

# Select 5-6 interesting countries based on your discovery
selected_countries <- c(
  # Your countries here - include top gainer, any decliner, and 3-4 others of interest
)

trajectory_data <- gapminder %>%
  # Your code here - filter for selected countries

# Create line plot
ggplot(trajectory_data, aes(x = year, y = lifeExp, color = country)) +
  # Your code here - add lines, points, and professional labeling

```

**Caption:** *\[Write 2-3 sentences explaining why you selected these countries and what pattern stands out\]*

# Research Discovery & Critical Analysis

## Choose Your Research Question

**Selected Research Question:** *\[A, B, C, D, or E\]*

*\[State your own research question here\]*

## Research Analysis

### Data Analysis

```{r}
#| label: research-analysis

# Your code here - use appropriate dplyr operations to investigate your question
# This should include filtering, grouping, summarizing, and/or calculating as needed

```

### Visualization

```{r}
#| label: research-plot
#| fig-width: 8
#| fig-height: 6

# Your code here - create one clear plot that supports your findings

```

### Interpretation (150-200 words)

*\[Your interpretation here - explain what you found, why it's interesting, potential explanations for the pattern, and any limitations of your analysis\]*
